Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}

### 2014-12-27 inst/doc/article.Rnw
version 0.9-1

<<echo=FALSE, results=hide>>=
# This should be set to FALSE in case you want to redo the sampling as well.
# Default is TRUE to save CRAN some time.
usePreCalcResults <- TRUE
@

@@

\documentclass[article, nojss]{jss}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

@@

%% almost as usual
\author{Gregor Kastner\\WU Vienna University of Economics and Business}
%\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol} \\ (DRAFT VERSION)}
\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}}
%\Shorttitle{The \proglang{R} package \pkg{stochvol} (DRAFT VERSION)} %% a short title (if necessary)
\Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting
%\Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol (DRAFT VERSION)} %% without formatting

@@

%% an abstract and keywords
\Abstract{
Heteroskedasticity in financial and economic datasets is a commonly observed feature, and the need to model this feature properly has for a long time been of great interest to researchers and practitioners alike. The \proglang{R} package \pkg{stochvol} provides a fully Bayesian implementation of heteroskedasticity modelling by means of the \emph{stochastic volatility} framework. Being a state-space formulation, it describes contemporaneous volatilities as latent random variables as opposed to deterministic values. The software described in this paper utilizes Markov chain Monte Carlo (MCMC) samplers to conduct inference by obtaining draws from the posterior distribution of parameters and latent variables, which can then be used for predicting future volatilities. The package can straightforwardly be employed as a stand-alone tool; moreover, it allows for easy incorporation into other MCMC samplers. Main focus of the paper is to show the functionality of \pkg{stochvol}, nevertheless it also provides a brief mathematical description of the model, an overview of the sampling schemes used, and an in-depth example using exchange rate data.
}
\Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
\Plainkeywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
%% at least one keyword must be supplied

@@

\section{Introduction}
When analyzing (financial) returns, focus is often laid on estimating and predicting potentially time varying volatilities. This interest has a long history, dating at least back to \cite{mar:por}, who investigated portfolio construction with optimal expected return-variance trade-off. In his article, he proposes rolling-window-type estimates for the instantaneous volatilities, but already then recognizes the potential for ``better methods, which take into account more information''.

One way of doing so is to model the evolution of volatility deterministically, i.e., through the (G)ARCH class of models. After the groundbreaking papers of \cite{eng:aut} and \cite{bol:gen}, these models have been generalized in numerous ways and were applied to a vast amount of real-world problems. As an alternative, \cite{tay:fin} proposes in his seminal work to model the volatility probabilistically, i.e., through a state-space model where the logarithm of the squared volatilities -- the latent states -- follow an autoregressive process of order one. Over time, this specification became known as the \emph{stochastic volatility (SV)} model. Even though several papers \citep[e.g.][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} find early evidence in favor of using SV, these models have found comparably little use in applied work. Reviewing this discrepancy, \cite{bos:rel} states:
\begin{quote}
 While there are literally thousands of applications of GARCH, for SV, this number is far lower. Two reasons for this relative lack of applied work using SV are apparent. First, there are as of yet no standard packages for estimating SV models, whereas for GARCH, most statistical packages have a wealth of options for incorporating GARCH effects. A second difference seems to be that GARCH has many variants of the model \citep{bol:glo}, with basically a single estimation method for all of them. For SV, there are few variants of the model, but a full series of estimation methods.
\end{quote}
In \cite{kas-fru:anc}, the latter issue has been thoroughly investigated, and an efficient MCMC estimation scheme
%using techniques such as sampling AWOL \citep[``all without a loop'', adapted from][]{rue:fas, mcc-etal:sim} and ASIS \citep[``ancillarity-sufficiency interweaving strategy'', adapted from][]{yu-men:cen}, an
has been proposed. The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} were crafted to cope with the first problem: the apparent lack of standard packages for efficiently estimating SV models.

@@

We briefly introduce the model and specify the notation used in the remainder of the paper. Furthermore, a quick overview over Bayesian parameter estimation via Markov chain Monte Carlo (MCMC) methods is given.

@@

Let $\yb = (y_1, y_2, \dots, y_T)^\top$ be a vector of returns with mean zero. The intrinsic feature of the SV model is that each observation $y_t$ is assumed to have its ``own'' contemporaneous variance $\e^{\hsv_t}$, thus relaxing the usual assumption of homoskedasticity. In order to make estimation of such a model feasible (note that there are just as many data points as there are variances!), this variance is not allowed to vary unrestrictedly with time. Rather, its logarithm is assumed to follow a parametric stochastic process, more specifically: an autoregressive process of order one. Note that feature is fundamentally different to GARCH-type models, where the time-varying volatility is assumed to follow a deterministic rather than stochastic evolution.

The SV model (in its centered parameterization) is thus given through
\begin{eqnarray}  
 \label{c1}
 y_{t}|\hsv_t &\sim&  \Normal{0, \exp{\hsv_{t}}},%\error_{t},% \qquad \error_{t} \sim \Normal{0,1},
 \\
 \label{c2}
 \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapar^2},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
 \label{c3}
\hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapar^2/(1-\phi^2)},
\end{eqnarray}

where $\Normal{\mu, \sigmapar^2}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapar^2$. We will refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} $\mupar$, the \emph{persistence} $\phipar$, and the \emph{volatility} $\sigmapar$ of log-variance. The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{T})$ appearing in state equations (\ref{c2}) and (\ref{c3}) is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely: the log-variance process). Note that the initial state $\hsv_0$ is distributed according to the stationary distribution of the autoregressive process of order one.

@@

The level $\mupar \in \mathbb{R}$ is equipped with the usual normal prior $\mupar \sim \Normal{b_{\mupar}, B_{\mupar}}$. In practical applications, this prior is usually chosen to be rather uninformative, e.g., through setting $b_\mupar = 0$ and $B_\mupar \geq 100$ for daily log-returns. Our experience with empirical data is that the exact choice is usually not very influential; see also Section~\ref{priorconfig}.

For the persistence parameter $\phipar \in (-1,1)$, we choose $(\phi+1)/2 \sim \Betadis{a_0, b_0}$, implying 
\begin{eqnarray}
 \label{beta_transformed}
 p(\phi) = \frac{1}{2\Betafun{a_0,b_0}}\left ( \frac{1+\phi}{2}\right ) ^{a_0-1}\left ( \frac{1-\phi}{2}\right ) ^{b_0-1},
\end{eqnarray}
where $a_0$ and $b_0$ are positive hyperparameters and $\Betafun{x,y} = \int_0^1t^{x-1}(1-t)^{y-1}\,dt$ denotes the beta function.
Clearly, the support of this distribution is the unit interval $(-1, 1)$; thus, stationarity of the autoregressive volatility process is guaranteed. Its expected value and variance are given through the expressions
\begin{eqnarray*}
 E(\phipar) &=& \frac{2a_0}{a_0+b_0}-1,\\
 V(\phipar) &=& \frac{4a_0b_0}{(a_0+b_0)^2(a_0+b_0+1)}.
\end{eqnarray*}
This obviously implies that the prior expectation of $\phipar$ depends only on the ratio $a_0:b_0$. It is greater than zero if and only if $a_0 > b_0$ and smaller than zero if and only if $a_0 < b_0$. For a fixed ratio $a_0:b_0$, the prior variance decreases with larger values of $a_0$ and $b_0$. The uniform distribution on the unit interval arises as a special case when $a_0=b_0=1$. For financial datasets with not too many observations (i.e., $T \lesssim 1000$), the choice of the hyperparameters $a_0$ and $b_0$ can be quite influential on the shape of the posterior distribution of $\phipar$. In fact, note that in the case when the underlying data-generating process is (near-)homoskedastic, the volatility of log-variance $\sigmapar$ will be (very close to) zero and thus the likelihood will contain little to no information about $\phipar$. Consequently, the posterior distribution of $\phipar$ will be (almost) equal to its prior, no matter how many data points are observed. For some discussion about this issue, see e.g., \cite{kim-etal:sto}, who choose $a_0=20$ and $b_0=1.5$, implying a prior mean of $0.86$ with a prior standard deviation of $0.11$ and thus very little mass for nonpositive values of $\phi$.

For the volatility of log-variance $\sigmapar \in \mathbb{R}^+$, we choose $\sigmapar^2 \sim B_{\sigmapar}\times \chi^2_1=\Gammad{1/2,1/2B_{\sigmapar}}$, which is motivated by \cite{fru-wag:sto}, who equivalently stipulate the prior for $\pm \sqrt { \sigmapar^2}$ to follow a centered normal distribution, i.e., $\pm \sqrt { \sigmapar^2} \sim \Normal{0,B_{\sigmapar}}$. As opposed to the usual Inverse-Gamma prior, this prior is not conjugate in the usual sampling scheme, however, does not bound $\sigmapar$ away from zero a priori. The choice of the hyperparameter $B_{\sigmapar}$ turns out to be of minor influence in empirical applications, as long as it is not picked to be too small.

@@

An MCMC algorithm such as the one implemented in the package \pkg{stochvol} will provide its user with draws from the posterior distribution of the desired random variables: in our case the latent log-variances $\hsvm$ and the parameter vector $\allpara$. Because these draws are usually dependent, Bayesian inference via MCMC may require careful design of the algorithm and attentive investigation of the draws obtained.

@@

In order to avoid the cost of code interpretation within each MCMC iteration, the core computations of the sampler are implemented in \proglang{C}, interfaced to \proglang{R} via the \pkg{Rcpp} package \citep{edd-fra:rcp}, where the convenience functions and the user-interface are implemented. This combination allows to make use of the well-established and widely accepted ease-of-use of \proglang{R} and its underlying functional programming paradigm. Moreover, existing frameworks for analyzing MCMC output such as \pkg{coda} \citep{plu-etal:cod} as well as high-level visualization tools can easily be used. Last but not least, users with a basic knowledge of \proglang{R} can use the package in a familiar surrounding with a very small entry cost. Nevertheless, despite all these convenience features, the package profits from highly optimized machine code generated by a compiler at package build time, thus providing acceptable runtime even for larger datasets.

A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of an ``ancillarity-sufficiency interweaving strategy'' (ASIS), which has been brought forward in the general context of state-space models by \cite{yu-men:cen}. ASIS exploits the fact that for certain parameter constellations, sampling efficiency improves substantially when considering a non-centered version of a state-space model.
This fact is commonly known as a reparameterization issue with an entire body of literature attached to it; for an early reference see e.g., \cite{hil-smi:par}.
In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility $\sigmapar$ of log-variance from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$. However, in the case of the SV model, it turns out that no single superior parameterization exists. Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better. To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered, and once in a noncentered parameterization. This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm. For more details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.

@@

The usual stand-alone approach to fitting SV models with \pkg{stochvol} follows the following workflow: (1) Prepare the data, (2) specify the prior distributions and configuration parameters, (3) run the sampler, (4) assess the output and display the results. All these steps will be described in more detail below, along with a worked example. For a stepwise incorporation of SV effects into other MCMC samplers, please see Section \ref{other}.

@@

The core sampling function \code{svsample} expects its input data \code{y} to be a numeric vector of returns without any missing values (\code{NA}s), and will throw an error if provided with anything else. In case that \code{y} contains zeros, a warning will be issued and a small offset constant of size \code{sd(y)/10000} will be added to the squared returns before doing the auxiliary mixture sampling \citep[cf.][]{omo-etal:sto}. A common and recommended way of avoiding zero returns is to de-mean the returns beforehand. Below is an example how to prepare data, illustrated with the \code{exrates} data set\footnote{The data set, which as been obtained from the European Central Bank's Statistical Data Warehouse, contains the daily bilateral prices of one Euro in 23 currencies from January 3, 2000, until April 4, 2012. Conversions to New Turkish Lira and Fourth Romanian Leu have been incorporated. See \code{?exrates} for more information.} which is included in the package. A visualization of one of these time series is displayed in Figure~\ref{fig1}.

<<usd1, fig=TRUE, include=FALSE, width=11, height=6>>=
set.seed(123)
library("stochvol")
data("exrates")
ret <- logret(exrates$USD, demean=TRUE)
par(mfrow=c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
plot(exrates$date, exrates$USD, type = 'l', main = "Price of 1 EUR in USD")
plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log-returns")
@

@@

Additionally to real-world data, \pkg{stochvol} has also a built-in data generator \code{svsim}. This function simply produces realizations of an SV process and returns an object of class \code{svsim}, which has its own \code{print}, \code{summary}, and \code{plot} methods. Exemplary code using \code{svsim} is given below, and the particular instance of this simulated series is displayed in Figure~\ref{fig2}.

<<usd2, fig=TRUE, include=FALSE, width=11, height=6>>=
sim <- svsim(500, mu = -9, phi = 0.99, sigma = 0.1)
par(mfrow=c(2, 1))
plot(sim)
@
\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd2}
 \caption{Visualization of a simulated time series as provided by the default \code{plot} method.}
 \label{fig2}
\end{center}
\end{figure}

@@

 After preparing the data vector \code{y}, the user needs to specify the prior hyperparameters for the parameter vector $\allpara = (\mupar, \phipar, \sigmapar)^\top$ -- see also Section~\ref{priors} -- and some configuration parameters. The appropriate values are passed to the main sampling function \code{svsample} as arguments, which are described below.

 The argument \code{priormu} is a vector of length $2$, containing mean and standard deviation of the normal prior for the level of the log-variance $\mupar$. A common strategy is to choose a vague prior here, e.g., \code{c(0, 100)}, because the likelihood usually carries enough information about this parameter. If one prefers to use (slightly) informative priors, e.g., to avoid outlier draws of $\mupar$, care must be taken whether actual log-returns or \emph{percentage} log-returns are analyzed. Assuming daily data, the former commonly have an unconditional variance of $0.0001$ or less and thus the level on the log-scale $\mupar$ lies around $\log(0.0001) \approx -9$, while the latter have the $100^2$-fold unconditional variance (around $1$), which implies a level of $\log(1) = 0$. Choices in the literature include \code{c(0, 10)} \citep{jac-etal:bayJE}, \code{c(0, 5)} \citep{yu:on}, \code{c(0, sqrt(10))} \citep{kim-etal:sto, mey-yu:bug} or \code{c(0, 1)} \citep{omo-etal:sto}. Note that most of these choices are quite informative and clearly designed for \emph{percentage} log-returns!

 For specifying the prior hyperparameters for the persistence of log-variance, $\phipar$, the argument \code{priorphi} may be used. It is again a vector of length $2$, containing $a_0$ and $b_0$ specified in Equation~\ref{beta_transformed}. As elaborated in Section~\ref{priors}, these values can possibly be quite influential, thus we advise to put some thought into choosing them and study the effect of different choices carefully. The default is currently given through \code{c(5, 1.5)}, implying a prior mean of $0.54$ and a prior standard deviation of $0.31$.

 The prior variance of log-variance hyperparameter $B_\sigmapar$ may be controlled through \code{priorsigma}. This argument defaults to \code{1} if not provided by the user. As discussed in Section~\ref{priors}, the exact choice of this value is usually not very influential in typical applications. In general, it should not be picked too small unless there is a very good reason, e.g., explicit prior knowledge, to do so.

For specifying the size of the burn-in, the parameter \code{burnin} is provided. It is the amount of MCMC iterations that are run but discarded to ensure convergence to the stationary distribution of the chain. The current default value for this parameter is $1000$, which has proved to suffice in most situations. Nevertheless, the user is encouraged to check convergence carefully, see Section \ref{check} for more details. The amount of iterations which are run after burn-in can be specified through the parameter \code{draws}, currently defaulting to $10\,000$. Consequently, the sampler will be run for a total of \code{burnin + draws} iterations.

>>>>

<<echo=FALSE, results=hide>>=
# This should be set to FALSE in case you want to redo the sampling as well.
# Default is TRUE to save CRAN some time.
usePreCalcResults <- TRUE

# Default here is FALSE because the object created is rather large
usePreCalcResultsExtra <- FALSE 
@

@@

\documentclass[article, nojss]{jss}
%\documentclass[article]{jss}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{textcomp}
\usepackage{tablefootnote}

@@

%% almost as usual
\author{Gregor Kastner\\WU Vienna University of Economics and Business}
\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}}
%\Shorttitle{The \proglang{R} package \pkg{stochvol} (DRAFT VERSION)} %% a short title (if necessary)
\Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting

@@

%% an abstract and keywords
\Abstract{
The \proglang{R} package \pkg{stochvol} provides a fully Bayesian implementation of heteroskedasticity modeling within the framework of stochastic volatility.
It utilizes Markov chain Monte Carlo (MCMC) samplers to conduct inference by obtaining draws from the posterior distribution of parameters and latent variables which can then be used for predicting future volatilities. The package can straightforwardly be employed as a stand-alone tool; moreover, it allows for easy incorporation into other MCMC samplers. The main focus of this paper is to show the functionality of \pkg{stochvol}. In addition, it provides a brief mathematical description of the model, an overview of the sampling schemes used, and several illustrative examples using exchange rate data.
}
\Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
\Plainkeywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
%% at least one keyword must be supplied

@@

\section{Introduction}
Returns -- in particular financial returns -- are commonly analyzed by estimating and predicting potentially time-varying volatilities. This focus has a long history, dating back at least to \cite{mar:por} who investigates portfolio construction with optimal expected return-variance trade-off. In his article, he proposes rolling-window-type estimates for the instantaneous volatilities, but already then recognizes the potential for ``better methods, which take into account more information''.

One approach is to model the evolution of volatility deterministically, i.e., through the (G)ARCH class of models. After the groundbreaking papers of \cite{eng:aut} and \cite{bol:gen}, these models have been generalized in numerous ways and applied to a vast amount of real-world problems. As an alternative, \cite{tay:fin} proposes in his seminal work to model the volatility probabilistically, i.e., through a state-space model where the logarithm of the squared volatilities -- the latent states -- follow an autoregressive process of order one. Over time, this specification became known as the \emph{stochastic volatility (SV)} model. Even though several papers \citep[e.g.,][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} provide early evidence in favor of using SV, these models have found comparably little use in applied work. This obvious discrepancy is discussed in \cite{bos:rel} who points out two reasons: the variety (and potential incompatibility) of estimation methods for SV models -- whereas the many variants of the GARCH model have basically a single estimation method -- and the lack of standard packages implementing these methods.

%While there are literally thousands of applications of GARCH, for SV, this number is far lower. Two reasons for this relative lack of applied work using SV are apparent. First, there are as of yet no standard packages for estimating SV models, whereas for GARCH, most statistical packages have a wealth of options for incorporating GARCH effects. A second difference seems to be that GARCH has many variants of the model \citep{bol:glo}, with basically a single estimation method for all of them. For SV, there are few variants of the model, but a full series of estimation methods.

 In \cite{kas-fru:anc}, the former issue is thoroughly investigated and an efficient MCMC estimation scheme
%using techniques such as sampling AWOL \citep[``all without a loop'', adapted from][]{rue:fas, mcc-etal:sim} and ASIS \citep[``ancillarity-sufficiency interweaving strategy'', adapted from][]{yu-men:cen}, an
is proposed. The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} are crafted to cope with the latter problem: the apparent lack of ready-to-use packages for efficiently estimating SV models.

@@

We begin by briefly introducing the model and specifying the notation used in the remainder of the paper. Furthermore, an overview of Bayesian parameter estimation via Markov chain Monte Carlo (MCMC) methods is given.

@@

Let $\yb = (y_1, y_2, \dots, y_n)^\top$ be a vector of returns with mean zero. The intrinsic feature of the SV model is that each observation $y_t$ is assumed to have its ``own'' contemporaneous variance $\e^{\hsv_t}$, thus relaxing the usual assumption of homoskedasticity. In order to make the estimation of such a model feasible, 
%(note that there are just as many data points as there are variances!)
this variance is not allowed to vary unrestrictedly with time. Rather, its logarithm is assumed to follow an autoregressive process of order one. Note that this feature is fundamentally different to GARCH-type models where the time-varying volatility is assumed to follow a deterministic instead of a stochastic evolution.

The SV model (in its centered parameterization) is thus given through
\begin{eqnarray}  
 \label{c1}
 y_{t}|\hsv_t &\sim&  \Normal{0, \exp{\hsv_{t}}},%\error_{t},% \qquad \error_{t} \sim \Normal{0,1},
 \\
 \label{c2}
 \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapartwo},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
 \label{c3}
\hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapartwo/(1-\phi^2)},
\end{eqnarray}

where $\Normal{\mu, \sigmapartwo}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapartwo$. We refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} of log-variance $\mupar$, the \emph{persistence} of log-variance $\phipar$, and the \emph{volatility} of log-variance $\sigmapar$. The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{n})$ appearing in state equations (\ref{c2}) and (\ref{c3}) is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely, the log-variance process). Note that the initial state $\hsv_0$ is distributed according to the stationary distribution of the autoregressive process of order one.

@@

The level $\mupar \in \mathbb{R}$ is equipped with the usual normal prior $\mupar \sim \Normal{b_{\mupar}, B_{\mupar}}$. In practical applications, this prior is usually chosen to be rather uninformative, e.g., through setting $b_\mupar = 0$ and $B_\mupar \geq 100$ for daily log returns. Our experience with empirical data is that the exact choice is usually not very influential; see also Section~\ref{priorconfig}.

For the persistence parameter $\phipar \in (-1,1)$, we choose $(\phi+1)/2 \sim \Betadis{a_0, b_0}$, implying 
\begin{eqnarray}
 \label{beta_transformed}
 p(\phi) = \frac{1}{2\Betafun{a_0,b_0}}\left ( \frac{1+\phi}{2}\right ) ^{a_0-1}\left ( \frac{1-\phi}{2}\right ) ^{b_0-1},
\end{eqnarray}
where $a_0$ and $b_0$ are positive hyperparameters and $\Betafun{x,y} = \int_0^1t^{x-1}(1-t)^{y-1}\,dt$ denotes the beta function.
Clearly, the support of this distribution is the interval $(-1, 1)$; thus, stationarity of the autoregressive volatility process is guaranteed. Its expected value and variance are given through the expressions
\begin{eqnarray*}
 E(\phipar) &=& \frac{2a_0}{a_0+b_0}-1,\\
 V(\phipar) &=& \frac{4a_0b_0}{(a_0+b_0)^2(a_0+b_0+1)}.
\end{eqnarray*}
This obviously implies that the prior expectation of $\phipar$ depends only on the ratio $a_0:b_0$. It is greater than zero if and only if $a_0 > b_0$ and smaller than zero if and only if $a_0 < b_0$. For a fixed ratio $a_0:b_0$, the prior variance decreases with larger values of $a_0$ and $b_0$. The uniform distribution on $(-1,1)$ arises as a special case when $a_0=b_0=1$. For financial datasets with not too many observations (i.e., $n \lesssim 1000$), the choice of the hyperparameters $a_0$ and $b_0$ can be quite influential on the shape of the posterior distribution of $\phipar$. In fact, when the underlying data-generating process is (near-)homoskedastic, the volatility of log-variance $\sigmapar$ is (very close to) zero and thus the likelihood contains little to no information about $\phipar$. Consequently, the posterior distribution of $\phipar$ is (almost) equal to its prior, no matter how many data points are observed. For some discussion about this issue, see, e.g., \cite{kim-etal:sto} who choose $a_0=20$ and $b_0=1.5$, implying a prior mean of $0.86$ with a prior standard deviation of $0.11$ and thus very little mass for nonpositive values of $\phi$.

For the volatility of log-variance $\sigmapar \in \mathbb{R}^+$, we choose $\sigmapartwo \sim B_{\sigmapar}\times \chi^2_1=\Gammad{1/2,1/2B_{\sigmapar}}$. This choice is motivated by \cite{fru-wag:sto} who equivalently stipulate the prior for $\pm \sqrt { \sigmapartwo}$ to follow a centered normal distribution, i.e., $\pm \sqrt { \sigmapartwo} \sim \Normal{0,B_{\sigmapar}}$. As opposed to the more common Inverse-Gamma prior for $\sigmapartwo$, this prior is not conjugate in the usual sampling scheme. However, it does not bound $\sigmapartwo$ away from zero a priori. The choice of the hyperparameter $B_{\sigmapar}$ turns out to be of minor influence in empirical applications as long as it is not set too small.

@@

An MCMC algorithm such as the one implemented in the package \pkg{stochvol} provides its user with draws from the posterior distribution of the desired random variables; in our case with the latent log-variances $\hsvm$ and the parameter vector $\allpara$. Because these draws are usually dependent, Bayesian inference via MCMC may require careful design of the algorithm and attentive investigation of the draws obtained.

@@

In order to avoid the cost of code interpretation within each MCMC iteration, the core computations are implemented in \proglang{C}. Their output is interfaced to \proglang{R} via the \pkg{Rcpp} package \citep{edd-fra:rcp}; there, the convenience functions and the user interface are implemented. This combination allows to make use of the well-established and widely accepted ease-of-use of \proglang{R} and its underlying functional programming paradigm. Moreover, existing frameworks for analyzing MCMC output such as \pkg{coda} \citep{plu-etal:cod} as well as high-level visualization tools can easily be used. Last but not least, users with a basic knowledge of \proglang{R} can use the package with a very low entry cost. Nevertheless, despite all these convenience features, the package profits from a highly optimized machine code generated by a compiler at package build time, thus providing acceptable runtime even for larger datasets.

A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of a variant of the ``ancillarity-sufficiency interweaving strategy'' (ASIS) which has been brought forward in the general context of state-space models by \cite{yu-men:cen}. ASIS exploits the fact that for certain parameter constellations sampling efficiency improves substantially when considering a non-centered version of a state-space model.
This is commonly referred to as a reparameterization issue and dealt with extensively in scholarly literature; for an early reference see, e.g., \cite{hil-smi:par}.
In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility of log-variance $\sigmapar$ from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$. However, in the case of the SV model, it turns out that no single superior parameterization exists. Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better. To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered and once in a noncentered parameterization. This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm. For more details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.

@@

The usual stand-alone approach to fitting SV models with \pkg{stochvol} exhibits the following workflow: (1) Prepare the data, (2) specify the prior distributions and configuration parameters, (3) run the sampler, (4) assess the output and display the results. All these steps are described in more detail below, along with a worked example. For a stepwise incorporation of SV effects into other MCMC samplers, see Section \ref{other}.

@@

The core sampling function \code{svsample} expects its input data \code{y} to be a numeric vector of returns without any missing values (\code{NA}s) and throws an error if provided with anything else. In the case that \code{y} contains zeros, a warning is issued and a small offset constant of size \code{sd(y)/10000} is added to the squared returns before doing the auxiliary mixture sampling \citep[cf.][]{omo-etal:sto}. However, we generally recommend to avoid zero returns altogether, e.g., by demeaning them beforehand.

Below is an illustration of how to prepare data by using the sample dataset \code{exrates}\footnote{The dataset -- obtained from the European Central Bank's Statistical Data Warehouse -- contains the daily bilateral prices of one euro in 23 currencies from January 3, 2000 until April 4, 2012. Conversions to New Turkish lira and Fourth Romanian leu have been incorporated. See \code{?exrates} for more information.} included in the package. Figure~\ref{fig1} provides a visualization of a time series from this dataset.

<<usd1, fig=TRUE, include=FALSE, width=11, height=6>>=
set.seed(123)
library("stochvol")
data("exrates")
ret <- logret(exrates$USD, demean = TRUE)
par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
plot(exrates$date, exrates$USD, type = 'l', main = "Price of 1 EUR in USD")
plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log returns")
@

@@

Additionally to real-world data, \pkg{stochvol} also has a built-in data generator \code{svsim}. This function simply produces realizations of an SV process and returns an object of class \code{svsim} which has its own \code{print}, \code{summary}, and \code{plot} methods. Exemplary code using \code{svsim} is given below and the particular instance of this simulated series is displayed in Figure~\ref{fig2}.

<<usd2, fig=TRUE, include=FALSE, width=11, height=6>>=
sim <- svsim(500, mu = -9, phi = 0.99, sigma = 0.1)
par(mfrow = c(2, 1))
plot(sim)
@
\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd2}
 \caption{Visualization of a simulated time series as provided by the default \code{plot} method.}
 \label{fig2}
\end{center}
\end{figure}

@@

 After preparing the data vector \code{y}, the user needs to specify the prior hyperparameters for the parameter vector $\allpara = (\mupar, \phipar, \sigmapar)^\top$ -- see also Section~\ref{priors} -- and some configuration parameters. The appropriate values are passed to the main sampling function \code{svsample} as arguments which are described below.

 The argument \code{priormu} is a vector of length $2$, containing mean and standard deviation of the normal prior for the level of the log-variance $\mupar$. A common strategy is to choose a vague prior here, e.g., \code{c(0, 100)}, because the likelihood usually carries enough information about this parameter. If one prefers to use (slightly) informative priors, e.g., to avoid outlier draws of $\mupar$, one must pay attention to whether \emph{log returns} or \emph{percentage log returns} are analyzed. Assuming daily data, log returns commonly have an unconditional variance of $0.0001$ or less and thus the level on the log scale $\mupar$ lies around $\log(0.0001) \approx -9$. Percentage log returns, on the other hand, have the $100^2$-fold unconditional variance (around $1$) which implies a level of $\log(1) = 0$. Choices in the literature include \code{c(0, 10)} \citep{jac-etal:bayJE}, \code{c(0, 5)} \citep{yu:on}, \code{c(0, sqrt(10))} \citep{kim-etal:sto, mey-yu:bug} or \code{c(0, 1)} \citep{omo-etal:sto}. Note that most of these choices are quite informative and clearly designed for percentage log returns.

 For specifying the prior hyperparameters for the persistence of log-variance, $\phipar$, the argument \code{priorphi} may be used. It is again a vector of length $2$, containing $a_0$ and $b_0$ specified in Equation~\ref{beta_transformed}. As elaborated in Section~\ref{priors}, these values can possibly be quite influential, thus we advise to choose them carefully and study the effects of different choices. The default is currently given through \code{c(5, 1.5)}, implying a prior mean of $0.54$ and a prior standard deviation of $0.31$.

 The prior variance of log-variance hyperparameter $B_\sigmapar$ may be controlled through \code{priorsigma}. This argument defaults to \code{1} if not provided by the user. As discussed in Section~\ref{priors}, the exact choice of this value is usually not very influential in typical applications. In general, it should not be set too small unless there is a very good reason, e.g., explicit prior knowledge, to do so.

For specifying the size of the burn-in, the parameter \code{burnin} is provided. It is the amount of MCMC iterations that are run but discarded to ensure convergence to the stationary distribution of the chain. The current default value for this parameter is $1000$ which has turned out to suffice in most situations. Nevertheless, the user is encouraged to check convergence carefully; see Section \ref{check} for more details. The amount of iterations which are run after burn-in can be specified through the parameter \code{draws}, currently defaulting to $10\,000$. Consequently, the sampler is run for a total of \code{burnin + draws} iterations.

 Three thinning parameters are available which all are $1$ if not specified otherwise. The first one, \code{thinpara}, is the denominator in the fraction of parameter draws (i.e., draws of $\allpara$) that are stored. E.g., if \code{thinpara} equals $10$, every $10$th draw is kept. The default parameter thinning value of $1$ means that all draws are saved. The second thinning parameter, \code{thinlatent}, acts in the same way for the latent variables $\hsvm$. The third thinning parameter, \code{thintime}, refers to thinning with respect to the time dimension of the latent volatility.
In the case that \code{thintime} is greater than 1, not all elements of $\hsvm$ are stored, e.g., for \code{thintime} equaling $10$, only the draws of $\hsv_1, \hsv_{11}, \hsv_{21}, \dots$ (and $\hsv_0$) are kept.
